---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Overview

<!-- badges: start -->
<!-- badges: end -->

**gtapshape** is an R package designed for the flexible aggregation of
land use and land cover data specifically tailored for use with the
[GTAP Agro-Ecological Zones (AEZ) model](https://www.gtap.agecon.purdue.edu/resources/res_display.asp?RecordID=2605). This
package provides tools to efficiently manipulate and aggregate spatial
and country-level data on global land use and land cover to the GTAP
regions and sectors.


## Installation

You can install the latest version of `gtapshape` from
[GitHub](https://github.com/) with:

```{r eval=FALSE}
## Install devtools if you don't already have it
install.packages("devtools")
## Install gtapshapeagg from GitHub
devtools::install_github("nvilloria/gtapshape", build_vignettes = TRUE)
```

## Usage

The function `build.dbase.from.sf` processes all the data needed to
create the land use and land cover headers and sets needed to split
national land rents into subnational land rents. The function takes
four inputs: 
* An R special features file with the subnational
boundaries (SF files for the 18 standard agroecological zones used in
the GTAP-AEZ model and the [14 WWF biomes](https://en.wikipedia.org/wiki/List_of_terrestrial_ecoregions_(WWF))
are included in the package), 
* The year for which the FAO data on production, prices and area harvested should
be processed (2011-2022 are included in the package).
* The source of the gridded data on global crop production (either
[Monfreda et al.](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2007GB002947)
or [CROPGRIDS](https://www.nature.com/articles/s41597-024-03247-7),
both provided in the package).
* The name of the har file that will be generated with the
physical data on land use and land cover needed by the GTAP AEZ
mode. 

The processed data are stored in a
[har](https://www.copsmodels.com/webhelp/rungtap/index.html?hc_harfiles.htm)
file with eight headers:

* QCR8: Crop production (MT) for the 8 gtap crops'
* VCR8: Value of crop production (1000 USD) for the 8 gtap crops'
* HARV: Harvested area (ha) for the 8 gtap crops'
* $QLV3: Livestock production (heads) for the 3 gtap livestock sectors'
* VLV3: Livestock output value (1000 USD) for the 3 gtap livestock sectors'
* LAND: Land cover areas (ha)'
* RTMB: Timber land rents (USD Million)'

In addition, `build.dbase.from.sf` writes a har file with the
following headers (sets):

* REG: GTAP regions (defaults to the GTAP database V11c).
* SUBN: Subnational boundaries (defaults to 18 AEZs)
* CRP8: The eight GTAP crop categories (pdr, wht, gro, v_f, osd, c_b, pfb, and ocr)
* CRP9: CRP8 + forest products (frs)
* LCOV: Seven land cover categoris (Forest, SavnGrasslnd, Shrubland,
  Cropland, Pastureland, Builtupland, Otherland)

Finally, `build.dbase.from.sf` writes a GTAP Aggregation Template text
file based on the GTAP Database V11c, that can be used to aggregate
the GTAP database.

The function defaults to splitting the data into 18 AEZs based on production,
area, and price data for year 2017, using the [Monfreda et al. (2008)](https://doi.org/10.1029/2007GB002947) data on
global gridded production, and produces three data files with the default names:
'gtaplulc.har', 'gtaplulc-sets.har', and 'gtaplulc-map.har':

```{r example, eval=FALSE}
library(gtapshape)
## Build the land use and land cover headers and sets needed to split
## subnational land rents into 18 AEZS:
aez18 <- build.dbase.from.sf()
```

The following call to the function updates the GTAP AEZ 18 database to
year 2022, using the more recent data on global gridded crop
production and area from CROPGRIDS (the resulting files are saved as
"gtaplulc-cropgrid-2022.har", "gtaplulc-cropgrid-2022-sets.har", and
"gtaplulc-cropgrid-2022-map.txt"):
```{r example2, eval=FALSE}
system.time(aez18_cropgrid_2022 <- build.dbase.from.sf(subnat_bound_file="aez18",
                                           crop_rasters = "cropgrids",
                                           year=2022,
                                           file="gtaplulc-cropgrid-2022.har"))

```

As another example, the user may choose a different scheme for the
subnational bounds, such as the 14 biomes defined by the WWF:

```{r example3, eval=FALSE}
system.time(biome14 <- build.dbase.from.sf(subnat_bound_file="biomes14",
                                           file="gtaplulc-biome14-2017.har"))
```

The user has flexibility in changing these defaults. Users with
knowledge of R can adapt the functions and workflow to their needs. 
Please refer to the following vignette for a more in-depth explanation 
of the functions built into the `build.dbase.from.sf` command:
```{r v1, eval=FALSE}
vignette('Country_level_landcover_ts.html')
```

In the companion package
[`gtapshapeagg`](https://github.com/nvilloria/gtapshapeagg) we provide
routines that split the national land rents of the standard GTAP
database using the sub-national data explained above. It also allows
for custom aggregations to different sectors and regions in a similar
fashion to
[FlexAgg2](https://www.gtap.agecon.purdue.edu/databases/flexagg2.asp).



## Data Preprocessing:

Detailed descriptions of the procedures used to process the different
land use and land cover datasets and generate the necessary inputs 
for the GTAP-AEZ framework are provided in the following vignettes:


### Building the shapefile with 18 AEZs:
Executing the following code will open the vignette which describes how we  
created the 18 Agro-Ecological Zones (AEZ) shapefile included in this package.
```{r v1, eval=FALSE}
vignette('create.18.aez.shapefile.html')
```

### Processing raster data on global land cover circa 2000:
(Micah)
```{r v1, eval=FALSE}
vignette('land.cover.html')
```

### Downloading and processing data from FAOSTAT:
(Micah)
```{r v1, eval=FALSE}
vignette('faostat.data.html')
```

### Updating land cover to specified year using FAOSTAT data:
(Micah)
```{r v1, eval=FALSE}
vignette('Country_level_landcover_ts.html')
```

### Processing crop and livestock production rasters:
(Micah)
[Monfreda, C., N. Ramankutty, and J. A. Foley (2008)](https://doi.org/10.1029/2007GB002947) 
[Robinson et al. (2014)](https://doi.org/10.1371/journal.pone.0096084)
```{r v1, eval=FALSE}
vignette('land.use.html')
```
[CROPGRIDS](https://doi.org/10.1038/s41597-024-03247-7)
```{r v1, eval=FALSE}
vignette('CROPGRIDS.convert.to.dataframes.html')
```

### Pre-processing forest rents
(Micah)
```{r v1, eval=FALSE}
vignette('preprocessing.of.forest.rents.html')
```


## Authors

- **Micah Cameron-Harp**
- **Nelson B. Villoria**

## Acknowledgments

- Thanks to the [GTAP](https://www.gtap.agecon.purdue.edu/) Center for
  developing and providing open source and free access to the the
  Agro-Ecological Zones data and modeling framework.

## Contact

For questions or feedback, please contact [Micah Cameron-Harp](mailto:mcameronharp@ksu.edu) or [Nelson B. Villoria](mailto:nvilloriap@ksu.edu).
