---
title: "Build a GTAP database with land use split in subnational bounds"
author: "Micah Cameron-Harp & Nelson B. Villoria"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a GTAP database with land use split in subnational bounds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval = FALSE}
devtools::document("../../gtapshape")
require(dplyr)
```

A number of functions carries the work of combining the time-invariant
gridded data on land use and land cover with the annual national data
published by FAOSTAT in order to get a database of land use and land
cover compatible with the theory behind the GTAP-AEZ model. These
functions are used many times. For example, the function
`all_uses_by_subnatbound()`, which allocates the gridded data to the
country and subnational boundaries of interest, e.g., 18
agroecological zones, is run 182 times (once for each of 172 crops, 7
land covers, and 3 livestock species), creating 182 dataframes. These
dataframes are then passed to `subnatbound.shares()` which estimates
the shares. The resulting 182 dataframes with subnational bounds are
then combined with the FAOSTAT data for each of the 182 uses
(`share.out.sectors.to.subnatbound()`). Further aggregation is
performed by additional functions, and finally, a har file with the
following six headers is written:

- Crop production (MT) for the 8 gtap crops
- Value of crop production (1000 USD) for the 8 gtap crops
- Harvested area (ha) for the 8 gtap crops
- Livestock production (heads) for the 3 gtap livestock sectors
- Livestock output value (1000 USD) for the 3 gtap livestock sectors
- Land cover areas (ha)

In principle, these functions can be used individually, and should
accomodate new sources of data with a modest programing effort. In
practice, most of the inputs to these functions remain fixed for most
applications. For cases in which the user only needs to change the
subnational boundaries, and or the base year of the database, we have
wrapped all the workflow connecting the individual functions in a
convenient wrapper function named `build.dbase.from.sf()`. The rest of
this vignette first discusses the workflow connecting the individual
functions, and then discuss the convenience wrapper
`build.dbase.from.sf()`.

# Building a land use GTAP-compatible land cover database from gridded and national data

```{r, eval = FALSE}

##  user  system elapsed 
## 24.51    2.92  103.28 
system.time(aez18 <- build.dbase.from.sf())
system.time(biomes14 <- build.dbase.from.sf(subnat_bound_file='biomes14', file="gtaplulc_biomes14.har"))

```

# Under the hood

## Inputs:

```{r, eval = FALSE}
## Make a gridded dataframe from the SF file with subnational
## boundaries:
g <- make_subnatbound_gridded_dataframe() ## Default is aez18
```
## Subnational bound shares of ag. production or and land cover used to share:

Gridded data on land use (crops annd livestock) and land cover are
aggregated to the subnational boundaries to calculate the national
shares of each land use or land cover that are then used to
dissagregate the country-level data from FAOSTAT to subnational
bounds. In what follows, The function `all_uses_by_subnatbound()`
passes the file names of the gridded data to
`aggregate_gridded_df_to_subnatbound()`, which aggregates them to the
subnational bounds.


```{r, eval = FALSE}
file.names.list <- list(
    crops = gridded.output.file.names <- list.files(system.file("monfreda", package = "gtapshape"),
                                            pattern = "\\Production.rda$", full.names = TRUE
                                            ),
    lstck = gridded.livestock.file.names <- c(
        system.file("fao_lstck_2005", "cattle.rda", package = "gtapshape"),
        system.file("fao_lstck_2005", "goats.rda", package = "gtapshape"),
        system.file("fao_lstck_2005", "sheep.rda", package = "gtapshape")),
    cover = gridded.landcover.file.names <- list.files(
        system.file("land_cover_2000", package = "gtapshape"),
        pattern = "\\.rda$", full.names = TRUE)
)

## Aggregate gridded land use and land cover to countries and subnational boundaries of interest

system.time( ## user: 78.25   system: 7.35  elapsed 85.84
    subnatbound.use.list <- lapply( file.names.list, function(.f){
        ha.by.iso.subnatbound.6covers <- all_uses_by_subnatbound(
            GADM_subnatbound_df= g,
            gridded.use.file.names= .f)
    }
    )
)
```
An important step in the construction of the shares is to map the
original uses in FAOSTAT into the GTAP sectors. The way in which this
aggregation needs to occur is specific to each type of data:

- For crops, the starting point are gridded data on crop production
  and harvested area fro the 172 crops covered by FAOSTAT. These data
  are taken from Monfreda et al. and are representative of year
  circa 2000.  These 172 crops are then aggregated to CPC sectors, so
  the subnational shares of each land use are at the CPC level. The
  country-level, annual FAOSTAT data is also aggregated to the CPC
  level before being shared out using the aggregated gridded
  data. After the annual, national FAOSTAT data has been assigned to
  each subnational bound, we aggregate them to the eight GTAP crop
  sectors.
- For livestock, both national (FAOSTAT) and gridded (Micah: Source)
  data are aggregated from three species (cattle, goats, and sheep)
  into the three livestock GTAP sectors: 'ctl', 'rmk', and 'wol'. As
  indicates in the `gridded.livestock.concordance`, this is a
  many-to-many mapping as the individual species are counted in more
  than one sector.
- For land covers, the national data (from FAOSTAT, updated as
  explained in the vignette where the updating process is explained)
  and the gridded data, from Ramankutty et al. (circa 2000), are given
  for the same number of sectors in the GTAP database, and therefore
  there is no need to map various sectors into aggregated GTAP
  categories as in the cases of crops and livestock. 
  
The following code first adds the mapping from 'uses' to the target
sectors (CPC sectors in the case of crops and GSC3 for livestock and
land covers) to the respective datasets, and then uses the function
`subnatbound.shares()` to calculate the subnational boundary shares of
the target sectors:

```{r, eval = FALSE}

## Add concordances to map FAOSTAST crops into CPC codes (aggregation
## to GSC3 sectors occurs after sharing out the national anual FAOSTAT
## data):
subnatbound.use.list[['crops']] <- merge(subnatbound.use.list[['crops']],
                                         unique(crop.concordance[, c("use", "item_code_cpc")]),
                                         by = "use",
                                         all.x = TRUE
                                         )

## Add concordances to map FAOSTAST livestock sectors to GSC3 sectors:
subnatbound.use.list[['lstck']] <- full_join(subnatbound.use.list[['lstck']],
                                            gridded.livestock.concordance,
                                            by = c("use"),
                                            relationship="many-to-many")


## Add concordances to rename land cover uses to the names use by the
## GTAP Center, labeled here as GSC3 sectors:
subnatbound.use.list[['cover']] <- merge(subnatbound.use.list[['cover']],
                                         land.cover.concordance,
                                         by = "use",
                                         all.x = TRUE)

system.time(
    shares <- lapply( subnatbound.use.list, function(.f){
        s <- subnatbound.shares(
            subnatbound.use= .f,
            sector=if(any(grepl("item_code_cpc", names(.f)))){"item_code_cpc"}else{"gsc3"}
        )
    }
    )
)
```

## Share out national data and aggregate to GTAP sectors:

The function `share.out.sectors.to.subnatbound()` combines the data at
the country level (stored below in the list object `iso.data`, with
attributes used to populate the HAR file written at the end) with the
subnational shares in order to distribute the national aggregates
across the subnational units. The data on crops requires a further
aggregation step, as the CPC sectors need to be aggregated to GSC3 sectors:

```{r, eval = FALSE}
## Annual national data, from FAOSTAT (4-letters data labels are used
## later for compatibility with the har file):
year = "2017"
iso.data <- list(
    QCR8 = iso.crop.production.2011.2022[
        iso.crop.production.2011.2022$year==year,],
    VCR8 = iso.crop.outputvalue.2011.2022[
        iso.crop.outputvalue.2011.2022$year==year,],
    HARV = iso.crop.harea.2011.2022[
        iso.crop.harea.2011.2022$year==year,],
    QLV3 = iso.gsc3lstk.heads.2011.2022[
        iso.gsc3lstk.heads.2011.2022$year==year,],
    VLV3 =iso.gsc3lstk.outputvalue.2011.2022[
        iso.gsc3lstk.outputvalue.2011.2022$year==year,],
    LAND = iso.land.cover.2011.2022[iso.land.cover.2011.2022$year==year,],
    RTMB = timber.rents.by.iso.2000
)

## Description of each dataset, used later when producing the har file:
attr(iso.data$QCR8, 'description')= 'Crop production (MT) for the 8 gtap crops'
attr(iso.data$VCR8, 'description')= 'Value of crop production (1000 USD) for the 8 gtap crops'
attr(iso.data$HARV, 'description')= 'Harvested area (ha) for the 8 gtap crops'
attr(iso.data$QLV3, 'description')= 'Livestock production (heads) for the 3 gtap livestock sectors'
attr(iso.data$VLV3, 'description')= 'Livestock output value (1000 USD) for the 3 gtap livestock sectors'
attr(iso.data$LAND, 'description')= 'Land cover areas (ha)'
attr(iso.data$RTMB, 'description')= 'Timber land rents (USD Million)'

## Dissagregates national data by subantional boundary and aggregates to GTAP regions
shares.o <- c(rep("crops",3), rep("lstck", 2), rep("cover",2) ) ## Index to select subnational boundaries shares
system.time( ## user: 5.27   system: 0.05   elapsed:  5.31
    gsc3.by.iso <- lapply(c(1:7), function(.n){
        ## .n <- 1
        .i <- iso.data[[.n]]
        .s <- shares.o[[.n]] 
        .l <- share.out.sectors.to.subnatbound(
            iso.data = .i,
            shares = shares[[.s]],
            units = unique(.i$unit),
            sector = if(any(grepl("item_code_cpc", names(.i)))){"item_code_cpc"}else{"gsc3"}
        )
        ## For the crop variables (production, output value and area
        ## harvested), further aggregation from CPC to GSC3 is needed:
        if( grepl('crops', .s ) ){
            ## Merge and aggregate
            merged <- merge(.l,
                        unique(crop.concordance[, c("item_code_cpc", "gsc3")]),
                        by = "item_code_cpc", all.x = TRUE
                        )
            result <- aggregate(value ~ reg + subnatbound + gsc3,
                                data = merged, sum, na.rm = TRUE)
            ## Add unit
            result$unit <- unique(.l$unit)
            ## Sort the result
            result <- result[order(result$reg, result$subnatbound, result$gsc3), ]
            rownames(result) <- NULL
            .l <- result
        }
        ## Assign attributes for har files:
        attr(.l,'description') <- attr(.i,'description')
        return(.l)
    }
    )
)

names(gsc3.by.iso) <- names(iso.data)
```

```{r, eval = FALSE}
lapply(gsc3.by.iso, function(x) attr(x,'description'))

## Verify number of regions, not tackled.
lapply( gsc3.by.iso, function(.x){
    length(unique(.x$reg))
    setdiff(regional.concordance$reg,unique(.x$reg))
})

lapply( gsc3.by.iso, function(.x){
    length(unique(.x$subnatbound))
})
```

## Order the data:
```{r, eval = FALSE}
#Use command 'order.gsc3.by.iso' to assign ordered levels to the character
#variables (like regions) so when they are output it is consistent with
#the GTAP base dataset
gsc3.by.iso.ordered <- order.gsc3.by.iso(gsc3.by.iso, 
                                         gtap_basedatasets_file = system.file("har", "gsdgset11cMV6.har", package = "gtapshape")
                                         )
```

## Write har file containing shared out land use and land cover data:
```{r, eval = FALSE}
write.gtaplulc.har(gsc3.by.iso.ordered, file="./gtaplulc.har")
```

## Write a har file containing the necessary sets:
```{r, eval = FALSE}
write.gtaplulcsets.har(gtap_basedatasets_file =
                           system.file("har", "gsdgset11cMV6.har", package = "gtapshape"),
                       file = './gtaplulcsets.har')
```

## Write aggregation file as .txt file
##    The subnat_bound_file should be the same as the one specified when constructing the LULC
##    database. If the user specifies "aez18" or "biomes14", the corresponding sf is lazyloaded
##    from the package 'data' folder.

```{r, eval = FALSE}
#Write aggregation file for 18 AEZs
write.gtaplulcagg.txt(subnat_bound_file="aez18",
                      base_aggr_file = system.file("mappings", 
                                                  "gtapv11c_aggr_base_file.txt",
                                                  package = "gtapshape"),
                      textfilename = 'gtaplulcagg.aez18.txt')

#Write aggregation file for 14 WWF Biomes
write.gtaplulcagg.txt(subnat_bound_file="biomes14",
                      base_aggr_file = system.file("mappings", 
                                                  "gtapv11c_aggr_base_file.txt",
                                                  package = "gtapshape"),
                      textfilename = 'gtaplulcagg.biomes14.txt')
```
