---
title: "Create 18 AEZ shapefile"
author: "Micah Cameron-Harp"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create 18 AEZ shapefile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette walks through the creation of the raster data depicting 18 Agro-Ecological Zones (AEZ) included in this package. To define the 18 AEZs, we follow the procedure used to create version 11 of the GTAP-AEZ Land Use and Land Cover database described in [Baldoz and Corong (2025)](https://www.gtap.agecon.purdue.edu/resources/res_display.asp?RecordID=7407). We use publicly available spatial data on the length of the growing period (LGP) and the thermal climate. Both of these datasets are available as rasters at the [GAEZ data viewer](https://gaez.fao.org/pages/data-viewer). We use the LGP and thermal zone data from 1981 to 2010 produced with the CRUTS32 Climate Model to create the 18 AEZ map in our package. The LGP raster we use contains the total number of growing period days, and the Thermal Zones raster from the Climate Classification sub-theme to determine thermal climates. Both of these rasters will be downloaded when the \link[gtapshape]{getrawdata} function is executed. Note, the same procedure we describe below could be used to create maps using more recent data.

```{r load, eval = FALSE}
#Load in required packages 
require(pacman)
pacman::p_load(
  terra,
  sf,
  dplyr)
```

## Length of growing period data

```{r, eval = FALSE}
#Load in length of growing season data
    lgd <- terra::rast("./raw_data/Total number of growing period days_1981 to 2010/lgd_CRUTS32_Hist_8110.tif")
    #Check RCS - WGS1984, EPSG 4326
    crs(lgd)  
    #Check resolution - It's 0.083 degrees which is equivalent to 5 arc minutes
    res(lgd) 
    #Check unique values
    unique(lgd)
    plot(lgd)
```



```{r, eval = FALSE} 
    #Recode so there are 36 categories defined by 10 degree intervals
      lgd_int_low <- c(-1, seq(from = 9, to = 349, by = 10))
      lgd_int_high <- c(seq(from = 9, to = 349, by = 10), 366)
      lgd_int_values <- c(seq(from = 1, to = 36, by = 1))
      lgd_int_df <- data.frame(lgd_int_low, lgd_int_high, lgd_int_values)
      lgd_mat <- data.matrix(lgd_int_df)
      lgd_36cat <- classify(lgd, lgd_mat)  
      unique(lgd_36cat)
      #Plot to check
      plot(lgd_36cat)
      #Check projection - WGS1984 - EPSG 4326 
      crs(lgd_36cat)  
      #Check resolution - 0.83 degrees = 5 arc minutes
      res(lgd_36cat) 

      #Clean up the environment
      rm(lgd, lgd_int_df, lgd_int_low, lgd_int_high, lgd_int_values)
      gc()
```

## Thermal zone data

```{r, eval = FALSE}
  #Thermal Zones
    mc2 <- terra::rast("./raw_data/Thermal zones/mc2_CRUTS32_Hist_8110.tif")
    #Make a key for the raster
    mc2_rast_values <- c(seq(from = 1, to = 13, by = 1))
    mc2_value_labels <- c("Tropics, warm", "Tropics, cool/cold/very cold", 
                          "Subtropics, warm/mod. cool", "Subtropics, cool", 
                          "Subtropics, cold", "Subtropics, very cold", 
                          "Temperate, cool", "Temperate, cold", 
                          "Temperate, very cold", "Boreal, cold", 
                          "Boreal, very cold", "Artic", "NA")
    mc2_label_key <- data.frame(mc2_rast_values, mc2_value_labels)

    #Reclassify Thermal Climates to only 3 categories
    mc2_3cat <- reclassify(mc2, c(0, 1, 1,
                                  1, 7, 2,
                                  7, 12, 3))  
    mc2_3cat_values <- c(1,2,3)
    mc2_3cat_labels <- c("Tropical", "Temperate", "Boreal")
    mc2_3cat_label_key <- data.frame(mc2_3cat_values, mc2_3cat_labels)
    mc2_3cat <- subs(mc2_3cat, mc2_3cat_label_key, by = 1, which = 2)
    #Plot to double check
    plot(mc2_3cat)
    #Check labels
    levels(mc2_3cat)
    #Check projection - WGS1984 - EPSG 9122 
    crs(mc2_3cat)  
    #Check resolution - 0.83 degrees = 5 arc minutes
    res(mc2_3cat) 
    
    #Clean up environment
    rm(mc2, mc2_rast_values, mc2_value_labels, mc2_label_key, mc2_3cat_values, mc2_3cat_labels, mc2_3cat_label_key)
    gc()
```

## Create 18 Agro-Ecological Zones

```{r setup,  eval = FALSE}

#Multiply the values for mc2_3cat by 100
  aez_108 <- mc2_3cat * 100 + lgd_36cat
  aez_cur_values <- c(unique(aez_108))
  aez_final_values <- c(seq(1, 108, by = 1))
  aez_key <- data.frame(aez_cur_values, aez_final_values)
  aez_mat <- data.matrix(aez_key)
  aez_108 <- reclassify(aez_108, aez_mat)    
  unique(aez_108)
  plot(aez_108)
  aez_108
  names(aez_108) <- c("aez_108")
  writeRaster(aez_108, filename = file.path(parent_dir, "Processed data/aez_108.tif"), overwrite = TRUE)

#Make another raster with the 18 aez categories instead
  aez_18_values <- c(rep(1:18, each = 6)) 
  aez_18_key <- data.frame(aez_final_values, aez_18_values)
  aez_18_mat <- data.matrix(aez_18_key)  
  aez_18 <- reclassify(aez_108, aez_18_mat)  
  unique(aez_18)  
  names(aez_18) <- c("aez_18")
  plot(aez_18)
```

## Output 18 AEZs as shapefile instead of raster

```{r, eval = FALSE}
#Convert raster to simple features (sf) object
  aez18_sf <- sf::st_as_sf(aez_18)
#Group geometries by AEZ and combine them to get one geometry per AEZ
  aez18_sf <- aez18_sf %>%
    dplyr::rename(aez = aez_18.tif)
  aez18_sf_out <- aez18_sf %>%
    group_by(aez) %>% 
    summarise(geometry = st_union(geometry))
#Make columns indicating the AEZ number
  aez18_sf_out <- aez18_sf_out %>%
    dplyr::mutate(subnat_num = aez,
                  subnat_name = paste0("AEZ", subnat_num)) %>%
    dplyr::select(subnat_name, subnat_num, geometry)
#Save to the gtapshape package data
  ##Uncomment the next line to overwrite the existing aez18 file included with the package
  #usethis::use_data(aez18, overwrite = TRUE)
```
