---
title: "CROPGRIDS Data (2020)"
author: "Micah Cameron-Harp and Nelson Villoria"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CROPGRIDS Data (2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

"CROPGRIDS is a comprehensive global, geo-referenced dataset providing area information for 173 crops circa the year 2020, at a resolution of 0.05˚ (~5.6 km at the equator). It represents a major update of the [Monfreda, C., N. Ramankutty, and J. A. Foley (2008)](https://doi.org/10.1029/2007GB002947) dataset (hereafter MRF), the most widely used geospatial dataset previously available, covering 175 crops with reference year 2000 at 10 km spatial resolution. [CROPGRIDS](https://doi.org/10.1038/s41597-024-03247-7) builds on information originally provided in MRF and expands it using 28 selected published gridded datasets, providing more recent crop information at regional, national, and global levels, largely covering the period 2015 – 2020. By combining the gridded information and the 2020 national-level statistics from FAOSTAT, CROPGRIDS reports the 2020 updated maps of harvested and crop (physical) area for 173 crops. The new CROPGRIDS data represent a useful input into modelling studies and sustainability analyses including in support of the 2030 Sustainable Development Agenda." (https://figshare.com/articles/dataset/CROPGRIDS/22491997)

```{r, eval = FALSE}
#Load in one crop to check .nc file types are effectively loaded by terra package.
  abaca_2020 <- terra::rast("./raw_data/CROPGRIDSv1.08_NC_maps/CROPGRIDSv1.08_abaca.nc")
  abaca_2020
```
There are multiple variables stored within the SpatRaster: harvested area, crop area, qual, set. We want harvested area. Note, the save command at the end of the following code block is commented out. It must be uncommented in order to overwrite the existing CROGRIDS .rda files included in the package. 

```{r, eval = FALSE}
#plot just the harvested area variable
  abaca_2020[names(abaca_2020)[2]]
  terra::plot(abaca_2020[[2]])
```
```{r, eval = FALSE}
#Load in 'gtapshape' commands
library(gtapshape)

#Now make a list of all the files, make sure they have some properties as current data, and save as dataframes in .rda format. 
  #List files
  CROPGRIDS.file.list <- list.files(path = "./raw_data/CROPGRIDSv1.08_NC_maps/",
                                     pattern = ".nc",
                                     full.names = TRUE)
  
  #Make sure we just do the ones for individual crops
  CROPGRIDS.file.list
  
  #Make a global raster with the intended CRS, resolution, and extent
  gr <- make_global_raster()

  #Use lapply to work through all of them
  lapply( CROPGRIDS.file.list, function(.r){
    # (.r <- CROPGRIDS.file.list[[1]]) #REMOVE BEFORE RUNNING FUNCTION
    r <- terra::rast(.r)
    r <- r[["harvarea"]]
    #Get the file name without the directory so we can store the crop name
    cropname <- gsub(pattern = 'CROPGRIDSv1.08_', replacement = '', basename(.r))
    cropname <- gsub(pattern = '.nc', replacement = '', cropname)
    names(r) <- cropname
    ## Ensure raster has the default resolution, extent, and
    ## coordinate reference system:
    r <- apply_global_raster_properties(input.raster = r, global.raster = gr)
    r <- as.data.frame(r, xy=TRUE)
    ## Keep only positive values to decrease file size:
    c <- colnames(r)[3]
    r1 <- r %>% dplyr::filter(!!sym(c)>0)
    ## Round up geographic coordinates to ensure compatibility with
    ## the country-geography raster used for aggregation:
    r1 <- round_up_coordinates(raster.df=r1)
    filename <- tools::file_path_sans_ext(basename(.r))
    assign(filename,r1)
    supplementary.data.path <- system.file("cropgrids", package = "gtapshape")
    ## Uncomment the next line to overwrite the existing cropgrids files included in the package
    # save(list=filename, file= file.path(supplementary.data.path, paste(filename,".rda",sep="")),
    #      compress = "xz",
    #      compression_level = 9)
}
)

##Clean-up environment
rm(list=ls()[! ls() %in% c("gr")])
gc()
```
