---
title: "Process global land cover rasters"
author: "Micah Cameron-Harp"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Process global land cover rasters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette will go over how we generate data on the 7 land cover types present in the GTAP Land Use and Land Cover datasets used in the *gtapshape* package. The data are created using publicly available raster data depicting the distribution of cropland and pastureland, potential vegetation classes, and urban areas. In each sub-section below, we detail how the underlying raster data was pre-processed. The raster data used in each sub-section are downloaded when the \link{getrawdata} function is executed. Note, some of the pre-processing described here mirrors the processes described by Baldos (2017) in "Development of GTAP version 9 Land Use and Land Cover database for years 2004, 2007 and 2011."	(Available at https://doi.org/10.21642/GTAP.RM30)

## **Urban land cover**
The urban land cover data are from Schneider, Friedl, and Potere (2010). The raster is categorical in nature, indicating the raster cells which are urban at a 500 meter resolution. A complete description of the data can be found in the article at: https://doi.org/10.1016/j.rse.2010.03.003. 
```{r, eval = FALSE}
  #Read in the urban land cover raster
    urban_rast <- terra::rast("urban_land_cover_modis/2002v5_urban_only_geog_mos.bip")
    #Check the header file
    hdr_file <- "urban_land_cover_modis/2002v5_urban_only_geog_mos.hdr"
    hdr_file <- readLines(con = hdr_file)
    #Display information in header file
    hdr_file
    #Correct the extent
    ext(urban_rast) <- c(-180, 180, -89.979167, 90)
```

We modify the base urban cover raster in two ways to ensure it is compatible with the other spatial data used in later processes. First, we reclassify the values of the raster so that urban areas have a value of 1 and all other areas have a value of zero.
```{r, eval = FALSE}
 #Reclassify so urban areas have a value of 1, all other areas are 0
    rcl_mat <- matrix(c(0, 12, 0,
                        12, 13, 1,
                        13, 255, 0), nrow = 3, ncol = 3, byrow = T)
    urban_rast <- terra::classify(urban_rast, rcl_mat, right = T)

  #Replace NA values with zeroes
    urban_rast <- terra::classify(urban_rast, cbind(NA, 0))
```

Second, we reproject the raster so it has the same Coordinate Reference System (EPSG 4326) and resolution (5 arc-minutes) as the other data. To do this, we use the \link{make.global.raster} and \link{apply_globa_raster_properties} functions.to make an empty raster with the target CRS and resolution. 
```{r, eval = FALSE}
  #Create a new global raster using the "make.global.raster" function with the
  #target CRS and resolution. We will use it to create a new urban land cover
  #raster with the same resolution and crs (projection).
    global_rast <- make.global.raster()
    #Project the urban cover raster into the new CRS and resample to correct resolution
    urban_rast <- apply_global_raster_properties(urban_rast, global_rast)
    #Save resulting output as .tif
    terra::writeRaster(urban_rast, 
                       filename = file.path(workdir_dir, 'workdir/rasters/urban_cover.tif'),
                       overwrite = TRUE)
    
  #Cleanup
    rm(list=ls())
    gc()
```

## **Potential vegetation**
The potential vegetation data are from Ramankutty and Foley (1999). The original raster data is categorical, displaying 13 different potential vegetation classes. A complete description of the data can be found in the article at: https://doi.org/10.1029/1999GB900046. The specific raster used in this vignette was downloaded in ASCII format from:https://sage.nelson.wisc.edu/data-and-models/datasets/global-potential-vegetation-dataset/. 

For our analyses, we collapse the original 13 classes into 4 categories: "Forests", "Shrubland", "Savanna + Grassland", and "Other Lands" as in Baldos (2017).  
```{r, eval = FALSE}
#Load in the Potential Vegetation Data from Ramankutty and Foley
  pot_veg_rast <- terra::rast("glpotveg_5min_orig.asc")

#Now we need to reclassify them into four categories as in Baldoz (2009)
  rcl_mat <- matrix(c(0, 8, 1,
                      10, 12, 2,
                      8, 10, 3,
                      12, 15, 4), nrow = 4, ncol = 3, byrow = T)
  pot_veg_4cat <- terra::classify(pot_veg_rast, rcl_mat)
  #Label the values
  pot_veg_4cat_values <- c(1,2,3,4)
  pot_veg_4cat_labels <- c("Forests", "Shrubland", "Savanna + Grassland", "Other Lands")
  pot_veg_4cat <- terra::subst( pot_veg_4cat,  pot_veg_4cat_values, pot_veg_4cat_labels)
```

Next, we save each of these four potential vegetation classes as individual rasters. Along with the cropland, pastureland, and urban land cover types, these make up the 7 total land cover types present in a GTAP LULC database.
```{r, eval = FALSE}
  #First create a global raster with the desired CRS and resolution to ensure
  #the potential vegetation rasters will conform with other raster data. 
    global_rast <- make.global.raster()
  
  #Create a forest potential veg raster
    rcl_mat <- matrix(c(0, 1, 1,
                        1, 4, 0), nrow = 2, ncol = 3, byrow = T)
    forest_pot_veg <- terra::classify(pot_veg_4cat, rcl_mat, right = T)
    #Project the forest potential vegetation raster into the new CRS and 
    #resample to correct resolution
    forest_pot_veg <- apply_global_raster_properties(forest_pot_veg, global_rast)
    terra::writeRaster(forest_pot_veg, filename = file.path(workdir_dir, "workdir/rasters/forest_pot_veg.tif"), overwrite = TRUE)

  #Shrubland potential veg
    #Note, reclassification matrix terms represent the order of the levels
    #of the categorical raster. So the order is now Forest, Other lands, Savanna,
    #then Shrubland.
    rcl_mat <- matrix(c(0, 3, 0,
                        3, 4, 1), nrow = 2, ncol = 3, byrow = T)
    shrubland_pot_veg <- terra::classify(pot_veg_4cat, rcl_mat, right = T)
    #Project the shrubland potential vegetation raster into the new CRS and 
    #resample to correct resolution
    shrubland_pot_veg <- apply_global_raster_properties(shrubland_pot_veg, global_rast)
    terra::writeRaster(shrubland_pot_veg, 
                       filename = file.path(workdir_dir, "workdir/rasters/shrubland_pot_veg.tif"),
                       overwrite = TRUE)

  #Savanna and grassland potential veg
    rcl_mat <- matrix(c(0, 2, 0,
                        2, 3, 1,
                        3, 4, 0), nrow = 3, ncol = 3, byrow = T)
    savanna_grass_pot_veg <- terra::classify(pot_veg_4cat, rcl_mat, right = T)
    #Project the savanna/grassland potential vegetation raster into the new CRS and 
    #resample to correct resolution
    savanna_grass_pot_veg <- apply_global_raster_properties(savanna_grass_pot_veg, global_rast)
    terra::writeRaster(savanna_grass_pot_veg, 
                       filename = file.path(workdir_dir, "workdir/rasters/savanna_grass_pot_veg.tif"),
                       overwrite = TRUE)

  #Other vegetation types like tundra, bare rock, etc.
    rcl_mat <- matrix(c(0, 1, 0,
                        1, 2, 1,
                        2, 4, 0), nrow = 3, ncol = 3, byrow = T)
    other_pot_veg <- terra::classify(pot_veg_4cat, rcl_mat, right = T)
    #Project the raster for the "other" category of potential vegetation into the
    #new CRS and resample to correct resolution
    other_pot_veg <- apply_global_raster_properties(other_pot_veg, global_rast)
    terra::writeRaster(other_pot_veg,  filename = file.path(workdir_dir, "workdir/rasters/other_pot_veg.tif"), overwrite = TRUE)
    
  #Cleanup
    rm(list=ls())
    gc()
```

## **Cropland and pastureland**
The cropland and pastureland data are described in the paper by Ramankutty et al. (2008), titled "Farming the planet: 1. Geographic distribution of global agricultural lands in the year 2000." The paper can be found at: https://doi.org/10.1029/2007GB002952. The raster data are categorical in nature, with cell values indicating whether the fraction of cell area categorized as cropland or pastureland at a 5 minute resolution. The two rasters used in this vignette, one for cropland and the other for pastureland, were downloaded from http://www.earthstat.org/cropland-pasture-area-2000/. 

First, we load in the cropland and pastureland rasters. We then use the \link{make.global.raster} and \link{apply_global_raster_properties} to ensure the resulting data have the target CRS (EPSG 4326) and resolution. 
```{r, eval = FALSE}
 #Cropland and pasture coverage
    #First create a global raster with the desired CRS and resolution
      global_rast <- make.global.raster()

    #Load cropland raster
      cropland_rast <- terra::rast("C:/Users/Micah/Dropbox/CGE_BIOMES_Carbon_2024/GTAP_BIOMES/raw_data/CroplandPastureArea2000_Geotiff/Cropland2000_5m.tif")
      #Make sure the cropland raster CRS and resolution match other rasters
      cropland_rast <- apply_global_raster_properties(cropland_rast, global_rast)
    
    #Load pastureland raster
      pasture_rast <- terra::rast("C:/Users/Micah/Dropbox/CGE_BIOMES_Carbon_2024/GTAP_BIOMES/raw_data/CroplandPastureArea2000_Geotiff/Pasture2000_5m.tif")
      #Make sure the pastureland raster CRS and resolution match other rasters
      pasture_rast <- apply_global_raster_properties(pasture_rast, global_rast)
```

## **Combine land cover types, normalize, and output as dataframes**
The final step in this vignette is to combine the various land covers and ensure the sum of all land cover fractions for any individual cell does not exceed 100% of the cell area. As in Baldos (2017), we begin by combining the cropland, pastureland, and urban area rasters. We then normalize any cells where the sum of the the cropland, pastureland, and urban area fractions is greater than 1.  
```{r, eval = FALSE}
  #Urban - created by urban_land_cover function
    urban_rast <- terra::rast(file.path(workdir_dir, 'workdir/rasters/urban_cover.tif'))

  #Join the cropland, pastureland, and urban rasters together
    rast_stack <- c(cropland_rast, pasture_rast, urban_rast)
    #Now make a new raster adding the others together
    sum_covers <- rast_stack[[1]] + rast_stack[[2]] + rast_stack[[3]]
      #some values look like they're above one, so more than 100% landcover
      sum_covers_vals <- c(values(sum_covers))
      sum_covers_df <- as.data.frame(sum_covers_vals)
      sum_covers_df <- sum_covers_df %>% dplyr::mutate(vals = as.numeric(sum_covers_vals))
      sum_covers_df <- sum_covers_df %>% dplyr::filter(!is.na(vals))

  #Now we need to normalize the cells with total covers greater than 1. To do
  #this we create a new raster which we will divide each of the land cover
  #rasters by. 
    rm(cropland_rast, pasture_rast, urban_rast)
    gc()
    #Make reclassification matrix
    m <- c(-1, 1, 1)
    #Make values that aren't >1 equal 1. Then when we use the divide the individual
    #rasters by the resulting raster, all values below one stay the same and those above 1 are normalized
    rclmat <- matrix(m, ncol = 3, byrow = TRUE)
    sum_covers_over1 <- terra::classify(sum_covers, rclmat)

    #Make new normalized rasters for cropland, pastureland, and urban areas
    norm_crop <- rast_stack[[1]]/sum_covers_over1
    norm_pasture <- rast_stack[[2]]/sum_covers_over1
    norm_urban <- rast_stack[[3]]/sum_covers_over1

    #Replace NA values with 0's and save the new normalized rasters
    norm_crop <- terra::classify(norm_crop, cbind(NA, 0))
    terra::writeRaster(norm_crop, 
                       filename = file.path(workdir_dir, 'workdir/rasters/crop_cover_fraction.tif'),
                       overwrite = TRUE)
    norm_pasture <- terra::classify(norm_pasture, cbind(NA, 0))
    terra::writeRaster(norm_pasture,
                       filename = file.path(workdir_dir, 'workdir/rasters/pasture_cover_fraction.tif'),
                       overwrite = TRUE)
    norm_urban <- terra::classify(norm_urban, cbind(NA, 0))
    terra::writeRaster(norm_urban,
                       filename = file.path(workdir_dir, 'workdir/rasters/urban_cover_fraction.tif'),
                       overwrite = TRUE)
```

Now, for cells where the sum of the urban, cropland, and pastureland area fractions is less than one, we allocate the remaining area fraction to the potential vegetation class.
```{r, eval = FALSE}
  #Last step is to read in the four potential vegetation raster to fill in the missing areas
      forest_pot_veg <- terra::rast(file.path(workdir_dir, 'workdir/rasters/forest_pot_veg.tif'))
      shrubland_pot_veg <- terra::rast(file.path(workdir_dir, 'workdir/rasters/shrubland_pot_veg.tif'))
      savanna_grass_pot_veg <- terra::rast (file.path(workdir_dir, 'workdir/rasters/savanna_grass_pot_veg.tif'))
      other_pot_veg <- terra::rast(file.path(workdir_dir, 'workdir/rasters/other_pot_veg.tif'))
  
      #Make a raster indicating the cell are remaining after accounting for the 
      #cropland, pastureland, and urban coverage fractions. Then fill in those
      #areas with the potential vegetation types.
        sum_croppasturb_norm_rasters <- norm_crop + norm_pasture + norm_urban
       #Now make a raster showing where we need to fill in land cover using the
        #potential vegetation rasters.
        pot_veg_to_fill <- 1 - sum_croppasturb_norm_rasters
      #Make final forest cover
        final_forest_cov <- forest_pot_veg * pot_veg_to_fill
        final_forest_cov <- terra::classify(final_forest_cov, cbind(NA, 0))
        terra::writeRaster(final_forest_cov, 
                           filename = file.path(workdir_dir, 'workdir/rasters/forest_cover_fraction.tif'), 
                           overwrite = TRUE)
      #Make final shrubland cover
        final_shrubland_cover <- shrubland_pot_veg * pot_veg_to_fill
        final_shrubland_cover <- terra::classify(final_shrubland_cover, cbind(NA, 0))
        writeRaster(final_shrubland_cover, 
                    filename = file.path(workdir_dir, 'workdir/rasters/shrubland_cover_fraction.tif'), 
                    overwrite = TRUE)
      #Make final grassland cover
        final_grass_cover <- savanna_grass_pot_veg * pot_veg_to_fill
        final_grass_cover <- classify(final_grass_cover, cbind(NA, 0))
        terra::writeRaster(final_grass_cover, 
                           filename = file.path(workdir_dir, 'workdir/rasters/grassland_cover_fraction.tif'),
                           overwrite = TRUE)
      #Make final other cover
        final_other_cover <- other_pot_veg * pot_veg_to_fill
        final_other_cover <- terra::classify(final_other_cover, cbind(NA, 0))
        terra::writeRaster(final_other_cover, 
                           filename = file.path(workdir_dir, 'workdir/rasters/other_cover_fraction.tif'), 
                           overwrite = TRUE)
```

